% USERS_MANUAL_EN.tex — Xinu (RAZ Extended Edition) User's Manual (English)
% Build: lualatex USERS_MANUAL_EN.tex   (run twice for the ToC)
\documentclass[a4paper,11pt]{article}

\usepackage{fontspec}
\setmonofont{Menlo}[Scale=0.86]
\usepackage[a4paper,margin=2.2cm]{geometry}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{fancyvrb}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}
\usepackage{newunicodechar}
\usepackage[hidelinks]{hyperref}
\usepackage{titlesec}

\newunicodechar{→}{\textrightarrow}
\newunicodechar{←}{\textleftarrow}
\newunicodechar{↑}{\textuparrow}
\newunicodechar{↓}{\textdownarrow}
\newunicodechar{↔}{\ensuremath{\leftrightarrow}}
\newunicodechar{×}{\ensuremath{\times}}
\newunicodechar{…}{\dots}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{≥}{\ensuremath{\geq}}
\newunicodechar{≈}{\ensuremath{\approx}}
\newunicodechar{≒}{\ensuremath{\fallingdotseq}}
\newunicodechar{°}{\textdegree}
\newunicodechar{—}{\textemdash}

\definecolor{codebg}{gray}{0.96}
\DefineVerbatimEnvironment{code}{Verbatim}%
  {fontsize=\small,frame=single,framerule=0.3pt,rulecolor=\color{gray!60},%
   framesep=5pt,xleftmargin=2pt,xrightmargin=2pt}

\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\titlespacing*{\section}{0pt}{14pt}{6pt}
\renewcommand{\arraystretch}{1.25}

\title{\textbf{Xinu (RAZ Extended Edition)\\User's Manual}\\[3pt]
  \large XFS filesystem, \texttt{cc} C compiler, \texttt{abclc} actor language,
  \texttt{wm} window manager}
\author{}
\date{}

\begin{document}
\maketitle
\vspace{-2.0em}
\begin{center}\rule{0.9\textwidth}{0.4pt}\end{center}

This manual describes an extended Embedded Xinu that adds a hierarchical
filesystem \textbf{XFS} (on a RAM disk), a tiny built-in C compiler
\textbf{cc} and bytecode VM \textbf{a.out}, an actor-oriented small language
\textbf{ABCL/c+} with its C translator \textbf{abclc}, a window manager
\textbf{wm} (a framebuffer GUI on ARM/QEMU), and Unix-like shell extensions
(\texttt{ls} / \texttt{cd} / \texttt{cat} / \texttt{make} / \texttt{edit}, \dots).
All stock Embedded Xinu facilities (threads, semaphores, networking) remain
available.

{\small\tableofcontents}
\bigskip

%======================================================================
\section{Introduction}
Xinu is a small UNIX-like OS originally written by Douglas Comer for teaching.
This extended edition adds a self-contained workbench that can edit, compile,
and run C / ABCL source on the OS itself --- so the whole develop-build-run
loop closes inside QEMU (or on real hardware) without a host cross-compiler.

\begin{center}\footnotesize
\begin{tabular}{@{}l l p{6.4cm}@{}}
\toprule
\textbf{Platform} & \textbf{Use} & \textbf{Notes}\\
\midrule
\verb|arm-qemu| & QEMU \verb|-M versatilepb| & UART shell + framebuffer + WM auto-start\\
\verb|arm-rpi|  & Raspberry Pi hardware     & UART / HDMI / USB keyboard\\
\bottomrule
\end{tabular}
\end{center}
XFS / cc / abclc / make / edit behave identically on every platform; the
WM-related pieces (\verb|rotlines|, \verb|wm_line|, \dots) are active on
\verb|arm-qemu| only.

%======================================================================
\section{Booting}
\subsection{Build}
\begin{code}
cd compile
make PLATFORM=arm-qemu        # default; arm-rpi also works
\end{code}
On success this produces \verb|compile/xinu.boot|.

\subsection{Run under QEMU}
With \verb|qemu-system-arm| installed, the bundled wrapper scripts are easiest.

\textbf{Console-only mode (no framebuffer):}
\begin{code}
./compile/run-console.sh
\end{code}
The UART shell (\verb|xsh$|) is multiplexed onto the terminal; quit with
\textbf{Ctrl-A \textrightarrow{} x}. Env vars: \verb|XINU_MEM| (default
\verb|128M|), \verb|XINU_PLATFORM| (default \verb|arm-qemu|).

\textbf{Window mode (with the WM demo):}
\begin{code}
./compile/run-window.sh
\end{code}
The LCD framebuffer appears in a macOS Cocoa window; the UART shell runs in the
launching terminal (\verb|-serial mon:stdio| keeps the QEMU monitor alongside).
Quit by closing the window or \textbf{Ctrl-A \textrightarrow{} x}. The window
scales smoothly up to 4× by dragging a corner (zoom-to-fit).

\subsection{Boot sequence}
The \verb|main()| thread in \verb|system/main.c| does, in order:
\begin{enumerate}[leftmargin=1.6em,itemsep=1pt]
\item print the OS version and memory layout;
\item \verb|xfsBootstrap()| --- format \verb|RAMDISK0| as XFS and mount it at \verb|/|;
\item seed \verb|/home/hello.c|, \verb|/home/sum.c|,
  \verb|.../PingPong.abcl|, \verb|.../RotLines.abcl|;
\item open the network device (when \verb|NETHER| is enabled);
\item open CONSOLE / TTY1;
\item on \verb|arm-qemu| only, \verb|create()| the \verb|wm_main| thread;
\item start a shell thread per CONSOLE / TTY1.
\end{enumerate}
When boot finishes, the prompt \verb|xsh$| means you can type commands.

\subsection{Running on real hardware (Raspberry Pi 3 B+)}
You can run this workbench on a real Raspberry Pi 3 B+ (BCM2837), not just under
QEMU. Build for the dedicated \verb|arm-rpi3| platform and boot from a microSD
card.

\textbf{Build for the board:}
\begin{code}
cd compile
make PLATFORM=arm-rpi3        # -> compile/xinu.boot (~305 KB)
\end{code}
If you changed a shared header, force a libxc rebuild:
\verb|rm -f lib/libxc/*.o lib/libxc.a && make PLATFORM=arm-rpi3|.

\textbf{What the SD card needs.} The Pi 3 firmware boots from a FAT32 partition.
Put on it the Broadcom firmware (\verb|bootcode.bin|, \verb|start.elf|,
\verb|fixup.dat| --- copied from any Raspberry Pi OS card; not in this repo),
\verb|kernel.img| (your built \verb|xinu.boot|, renamed), and \verb|config.txt|
(from \verb|compile/sdcard-rpi3/config.txt|). Key \verb|config.txt| lines:
\begin{center}\footnotesize
\begin{tabular}{@{}l p{8.2cm}@{}}
\toprule \textbf{Line} & \textbf{Meaning}\\ \midrule
\verb|arm_64bit=0|           & boot 32-bit (AArch32)\\
\verb|kernel=kernel.img|     & kernel file name\\
\verb|kernel_address=0x8000| & load address (0x8000)\\
\verb|enable_uart=1|         & enable the UART\\
\verb|dtoverlay=disable-bt|  & route PL011 (UART0) to GPIO14/15 (disable Bluetooth)\\
\verb|init_uart_clock=48000000| & PL011 reference clock 48 MHz (115200 8N1)\\
\verb|hdmi_force_hotplug=1|  & force HDMI output even with no monitor detected\\
\bottomrule
\end{tabular}
\end{center}

\textbf{Make the card (Option A --- fresh FAT32):}
\begin{code}
diskutil list                                        # find the microSD diskN (NOT disk0!)
diskutil eraseDisk FAT32 XINU MBRFormat /dev/diskN   # erases the card
cp /Volumes/bootfs/{bootcode.bin,start.elf,fixup.dat} /Volumes/XINU/
cp compile/sdcard-rpi3/config.txt /Volumes/XINU/
cp compile/xinu.boot /Volumes/XINU/kernel.img
diskutil eject /Volumes/XINU
\end{code}
\textbf{Option B --- reuse a Raspberry Pi OS card (no format):} copy only
\verb|kernel.img| and our \verb|config.txt| onto its boot partition (firmware is
already there; our \verb|config.txt| sets \verb|kernel=kernel.img|).

\textbf{Serial-console wiring.} Use a 3.3 V USB-TTL adapter (do \emph{not}
connect its 5 V): GND \textrightarrow{} pin 6, adapter RX \textrightarrow{}
pin 8 (GPIO14/TXD), adapter TX \textrightarrow{} pin 10 (GPIO15/RXD). Open at
115200 8N1: \verb|screen /dev/tty.usbserial-XXXX 115200|.

\textbf{Boot.} Insert the SD and power on. After ~10--40 s the \verb|xsh$| prompt
appears on the serial line (and, on the gwm build, the HDMI desktop). All
workbench commands (\verb|cc|, \verb|abclc|, \verb|make|, \verb|run|, \dots) work
as under QEMU. For the Pi 3 B+ desktop / WiFi / HTTP features see \verb|README.md|
and \verb|docs/USERS-GUIDE-JA.md|.

\textbf{Updating the kernel without swapping the SD (\texttt{/kexec}).} If a
\verb|/kexec|-capable build is already running:
\begin{code}
curl --data-binary @compile/xinu.boot "http://192.168.3.50:8080/upload?dst=xinu.boot"
curl -X POST http://192.168.3.50:8080/kexec
sleep 25
curl http://192.168.3.50:8080/api/mmu     # responds if the new kernel booted
\end{code}
This is volatile --- a power cycle reboots from the SD again.

\subsection{Connecting to WiFi (Pi 3 B+)}
The \verb|arm-rpi3| build has a BCM43455 WiFi driver (over the Arasan SDIO). It
is \textbf{not} auto-connected at boot --- run \verb|wifi on| each time. From the
\verb|xsh$| shell:
\begin{center}\footnotesize
\begin{tabular}{@{}l p{9.4cm}@{}}
\toprule \textbf{Command} & \textbf{Description}\\ \midrule
\verb|wifi status| & show the current connection (SSID + IP/GW)\\
\verb|wifi on| & interactive: scan (~30 s; the radio drops briefly) \textrightarrow{} list APs \textrightarrow{} pick by number \textrightarrow{} password (blank = open) \textrightarrow{} join + DHCP; then start the ARP/ICMP responder so the IP is pingable\\
\verb|wifi on <ssid> [pass]| & non-interactive (no scan/prompt); for the \verb|/shell?cmd=| remote-login route (no stdin). Omitted/blank pass = open\\
\verb|wifi off| & disconnect (bring the BSS down)\\
\verb|wifi-invest| & when "connected but ping fails", restart the responder, send a gratuitous ARP, retry the gateway probe until it works (or report the likely cause)\\
\bottomrule
\end{tabular}
\end{center}
\begin{code}
xsh$ wifi on
Scanning for access points (~30s; the radio drops briefly)...
Available networks:
  1: MyHome-5G
  2: Buffalo-G-56A2
Select AP number (0 = cancel): 1
Password for "MyHome-5G" (blank = open): ********
Connecting to "MyHome-5G"...
WiFi: connected to "MyHome-5G"
  ip 192.168.3.52  gw 192.168.3.1
\end{code}
A scan briefly drops the link (the radio retunes all channels); reconnect if
needed. WiFi does not auto-reconnect after reboot. The same is available over
HTTP on the wired control plane: \verb|/api/wifi/scan|,
\verb|/api/wifi/join?ssid=S&pass=P|, \verb|/api/wifi/dhcp|,
\verb|/api/wifi/ping?ip=A.B.C.D|.

\subsection{Mesh networking: multiple Xinu nodes (MANET ad-hoc)}
Several Xinu boards can form a peer-to-peer mesh with \textbf{no access point},
using 802.11 IBSS (ad-hoc) mode plus on-demand AODV routing --- the same stack
used in the drone-HIL demo (Pi 4 + Pi 3 nodes). Each node joins the same ad-hoc
cell with a distinct node number:
\begin{code}
xsh$ wifi adhoc <cell-ssid> [channel] [node]
\end{code}
\begin{itemize}[leftmargin=1.4em,itemsep=1pt]
\item \verb|<cell-ssid>| --- the cell name; \textbf{all nodes must use the same
  name and channel}.
\item \verb|[channel]| --- 802.11 channel (default 6).
\item \verb|[node]| --- this node's number (default 2); it gets the static IP
  \verb|10.0.0.<node>/24|. \textbf{Give every node a distinct number.}
\end{itemize}
On success the node prints \verb|ad-hoc up, ip 10.0.0.<node> (responder running)|
and answers pings on \verb|10.0.0.<node>|. Example --- a 3-node cell on channel 6:
\begin{code}
xsh$ wifi adhoc mesh1 6 1      # node 1 -> 10.0.0.1
xsh$ wifi adhoc mesh1 6 2      # node 2 -> 10.0.0.2
xsh$ wifi adhoc mesh1 6 3      # node 3 -> 10.0.0.3
\end{code}
Nodes within radio range reach each other directly at \verb|10.0.0.x| (a Mac
joined to the same IBSS cell with a \verb|10.0.0.x/24| address can
\verb|ping 10.0.0.1|, etc.).

\textbf{Multi-hop (AODV).} When a destination is not in direct range, the minimal
AODV module (M13) discovers a route on demand: the originator broadcasts an RREQ
(UDP port 654), each in-range node relays it and records a reverse route, and the
destination answers with an RREP that installs the forward route
(\verb|route to 10.0.0.x via 10.0.0.y (k hop)|). Every node runs the AODV control
plane (relay + reply) automatically as part of its responder, so intermediate
nodes forward traffic for peers beyond direct range (route table: up to 16
entries). A node can also be brought up over the wired HTTP control plane:
\verb|curl "http://<node-ip>:8080/wifi-adhoc?ssid=mesh1&ch=6&n=2"|.

IBSS / ad-hoc is independent of the infrastructure \verb|wifi on| mode (run
\verb|wifi off| to leave an AP first); like AP mode it is not restored after
reboot --- re-run \verb|wifi adhoc| on each node.

%======================================================================
\section{Using the shell}
\subsection{Basics}
\begin{code}
command [args...]    [< input-redirect] [> output-redirect] [&]
\end{code}
\verb|&| runs in the background; \verb|<| \verb|>| redirect to/from I/O devices;
there is no completion or history.

\subsection{Command list (added/extended by this edition)}
\begin{center}\footnotesize
\begin{longtable}{@{}l l p{7.4cm}@{}}
\toprule
\textbf{Command} & \textbf{Usage} & \textbf{Description}\\
\midrule \endhead
\verb|pwd|   & \verb|pwd|              & Print the current directory\\
\verb|cd|    & \verb|cd [PATH]|        & Change directory (default \verb|/|)\\
\verb|ls|    & \verb|ls [-l] [PATH]|   & Directory listing; \verb|-l| shows type/size\\
\verb|cat|   & \verb|cat FILE...|      & Print file contents\\
\verb|mkdir| & \verb|mkdir DIR...|     & Create directories\\
\verb|rmdir| & \verb|rmdir DIR...|     & Remove empty directories\\
\verb|touch| & \verb|touch FILE...|    & Create empty file / update mtime\\
\verb|rm|    & \verb|rm FILE...|       & Remove files\\
\verb|cp|    & \verb|cp SRC DST|       & Copy (DST truncated)\\
\verb|mv|    & \verb|mv SRC DST|       & Rename / move\\
\verb|write| & \verb|write FILE TEXT...| & Join args with spaces + newline into FILE\\
\verb|edit|  & \verb|edit FILE|        & Built-in Emacs-like editor\\
\verb|mkfs|  & \verb|mkfs DEVICE [VOL]| & Format a block device as XFS\\
\verb|mount| & \verb|mount DEV MNT|    & Mount an XFS\\
\verb|umount|& \verb|umount MNT|       & Unmount an XFS\\
\verb|cc|    & \verb|cc SRC [-o OUT]|  & Compile C source to a.out\\
\verb|abclc| & \verb|abclc SRC.abcl [-o OUT]| & ABCL \textrightarrow{} C \textrightarrow{} a.out in one step\\
\verb|make|  & \verb|make [TARGET...]| & Run the Makefile in the current directory\\
\verb|run|   & \verb|run PATH|         & Execute an a.out\\
\verb|rotlines| & \verb|rotlines [FRAMES] [DEG/F]| & Draw rotating lines on the WM\\
\verb|clear| & \verb|clear|            & Clear the screen\\
\verb|sleep| & \verb|sleep N|          & Sleep N ms\\
\bottomrule
\end{longtable}
\end{center}

\textbf{Implicit \texttt{run}.} If you type a command the shell does not
recognise, it tries \verb|aoutRun(<that name>)| as a last resort (around
\verb|shell/shell.c:332|). So after \verb|cc hello.c -o hello| you can launch the
program directly by its path-like name (\verb|xsh$ hello| \textrightarrow{}
\verb|hello...|).

%======================================================================
\section{XFS --- the filesystem}
A 4 KB-block hierarchical filesystem (\verb|include/xfs.h|), created on a 16 MB
RAM disk (\verb|RAMDISK0|) at boot and mounted at \verb|/|.

\begin{center}\footnotesize
\begin{tabular}{@{}l l@{}}
\toprule \textbf{Item} & \textbf{Value}\\ \midrule
Block size & 4096 B\\
Max filename length & 56 B\\
inode size & 128 B\\
Direct blocks & 12 (= 48 KB)\\
Single indirect & 1024 blocks (≈ 4 MB)\\
Double indirect & yes\\
Magic & \verb|XFS!| (0x58465321)\\
\bottomrule
\end{tabular}
\end{center}

Directories store \verb|struct xdirent[]| (64 B/entry) inside the inode, read
with \verb|xfsReaddir()|. \verb|xfsChdir()| / \verb|xfsGetcwd()| keep a
\textbf{per-thread} absolute-path CWD (\verb|cd /home| in shell A does not affect
shell B). \verb|xfsBootstrap()| checks the first 64 KB of \verb|RAMDISK0| for the
XFS magic, runs \verb|xfsMkfs| if absent, and mounts at \verb|/|.

Main API from C:
\begin{code}
int fd = xfsOpen("/home/foo", XFS_O_RDWR | XFS_O_CREAT | XFS_O_TRUNC);
xfsWrite(fd, buf, n);  xfsRead(fd, buf, n);
xfsSeek(fd, 0, XFS_SEEK_SET);  xfsClose(fd);
xfsMkdir("/home/sub");  xfsRename("/home/old", "/home/new");  xfsUnlink("/home/foo");
\end{code}
\textbf{RAMDISK0 is RAM: files disappear when QEMU exits.} Back up anything you
want to keep on the host.

%======================================================================
\section{cc / run --- the C compiler and launcher}
\subsection{Supported C subset}
Documented at the top of \verb|system/cc.c|: one source file, one
\verb|int main(void)| plus argument-less helpers; declarations \verb|int x;| /
\verb|int x = expr;| only; statements \verb|if/else|, \verb|while|, \verb|for|,
\verb|return|, blocks, expression statements; expressions over integers,
strings, identifiers, \verb|+ - * / %|, the comparisons,
\verb|! && ||| (logical), unary \verb|-|, parentheses; only built-in calls;
\verb|#include| accepted and ignored. \textbf{Unsupported:} structs, pointers,
arrays, floating point, multiple files.

\subsection{Built-in functions}
\begin{center}\footnotesize
\begin{tabular}{@{}l l p{6.8cm}@{}}
\toprule \textbf{BI} & \textbf{Function} & \textbf{Use}\\ \midrule
0 & \verb|printf(fmt, ...)| & string / integer formatting\\
1 & \verb|puts(s)| & string + newline\\
2 & \verb|putchar(c)| & one character\\
3 & \verb|getchar()| & one character of input\\
4 & \verb|exit(code)| & terminate\\
5 & \verb|rgb(r,g,b)| & BGR565 16-bit colour\\
6 & \verb|wm_line(idx,x1,y1,x2,y2,color)| & set WM user-line slot (0..7)\\
7 & \verb|wm_render(on)| & enable/disable user-line rendering\\
8 & \verb|wm_clear()| & clear user lines\\
9 & \verb|sleep_ms(ms)| & sleep\\
10/11 & \verb|isin(deg)| / \verb|icos(deg)| & Q12 fixed point (4096 = 1.0)\\
12/13 & \verb|screen_w()| / \verb|screen_h()| & screen size\\
\bottomrule
\end{tabular}
\end{center}

\subsection{a.out format and shell usage}
\verb|struct aout_header| in \verb|include/aout.h|:
\begin{code}
+--------------------------+
| magic = "XAOU"           |
| version = 1              |
| code_size / const_size   |
| entry / nlocals          |
+--------------------------+
| bytecode body            |
| string constant pool     |
+--------------------------+
\end{code}
The VM is stack-based (256 entries) with one-byte opcodes (\verb|OP_PUSH_I32|,
\verb|OP_LOAD_LOC|, \verb|OP_ADD|, \verb|OP_JZ|, \verb|OP_CALL_BI|); see
\verb|aoutRun()| in \verb|system/aout.c|.
\begin{code}
xsh$ cc /home/hello.c -o /home/hello
xsh$ run /home/hello
hello...
xsh$ /home/hello              # implicit run also works
\end{code}

%======================================================================
\section{abclc --- the ABCL/c+ actor language}
ABCL/c+ is a small actor-oriented language ("classes + message sends"). Here it
is two-stage: \textbf{abclc translates to C, and cc compiles that C to bytecode}.
\begin{code}
class ClassName {
    var field1;
    method methodName(p1, p2) {
        field1 = p1 + 1;
        send other.methodName(arg1, arg2);
    }
}
main {
    new ClassName a;  new ClassName b;
    send a.methodName(b, 10);
}
\end{code}
Implicit identifiers \verb|self| (current actor id) and \verb|sender|; at most
one \verb|send| per method (tail-call style), so each actor processes one message
at a time; all values are \verb|int|-equivalent; limits: 4 classes, 16
methods/class, 8 fields, 8 args.
\begin{code}
xsh$ cd /home/abclcp/abclc
xsh$ abclc PingPong.abcl
abclc: translated PingPong.abcl -> PingPong.c
abclc: compiled PingPong.c -> PingPong
xsh$ run PingPong
\end{code}
With \verb|-o NAME|, both \verb|NAME.c| (intermediate C) and \verb|NAME| (a.out)
are written. The emitted C uses only cc's built-ins plus a few scheduling
helpers; one message runs per time-slice, and after a \verb|send| the message is
queued.

%======================================================================
\section{make --- the minimal build system}
\verb|shell/xsh_make.c|: if the CWD has no \verb|Makefile|, scan \verb|*.c| and
\verb|*.abcl| and auto-generate one; read it and build each target in
\verb|TARGETS = ...| (or declaration order); \verb|.c|\,\textrightarrow{}\,
\verb|cc SRC -o TARGET|, \verb|.abcl|\,\textrightarrow{}\,\verb|abclc SRC -o TARGET|.
\begin{code}
# comment
TARGETS = hello sum PingPong RotLines
hello:    hello.c
sum:      sum.c
PingPong: PingPong.abcl
RotLines: RotLines.abcl
\end{code}
Targets with no explicit rule are inferred by checking \verb|TARGET.c| then
\verb|TARGET.abcl|.
\begin{code}
xsh$ cd /home
xsh$ make
make: generated Makefile (2 .c, 0 .abcl)
    cc hello.c -o hello
    cc sum.c -o sum
make: built 2 target(s)
xsh$ ./hello
hello...
\end{code}

%======================================================================
\section{edit --- the built-in editor}
\verb|shell/xsh_edit.c| is an Emacs-like modeless editor (up to 400 lines × 256
chars). \verb|edit /home/hello.c|; a non-existent path opens a new buffer.
\begin{center}\footnotesize
\begin{tabular}{@{}l l@{}}
\toprule \textbf{Key} & \textbf{Action}\\ \midrule
\verb|C-f| / \verb|C-b| & forward / back one character\\
\verb|C-n| / \verb|C-p| & down / up one line\\
\verb|C-a| / \verb|C-e| & start / end of line\\
←↑→↓ & arrows (ESC[ sequences)\\
\verb|C-d| & delete at cursor\\
\verb|Backspace| & delete previous character\\
\verb|Enter| & insert newline\\
\verb|Tab| & two spaces\\
\verb|C-k| & kill to end of line\\
\verb|C-l| & redraw\\
\verb|C-x C-s| & save\\
\verb|C-x C-c| & save and exit\\
\verb|C-g| & exit without saving\\
\bottomrule
\end{tabular}
\end{center}
The status line shows \verb|Lrow:Ccol  status|.

%======================================================================
\section{wm --- the window manager}
\verb|apps/wm.c| drives, on QEMU's VersatilePB, the PL110 LCD controller
(\verb|0x10120000|, 16bpp BGR565) and the PL050 PS/2 mouse (\verb|0x10007000|)
directly, showing a desktop-like GUI on a 1024×1024 virtual screen (640×480
physical). With \verb|run-window.sh|, \verb|wm_main| starts at boot. It provides
a status bar, three draggable demo windows, a mouse cursor, a user draw hook
\verb|wm_set_user_render(...)|, and direct drawing APIs (\verb|put_pixel_pub|,
\verb|fill_rect_pub|, \verb|rect_outline_pub|, \verb|draw_string_pub|,
\verb|wm_draw_line|).

From C / ABCL the basic pattern is \verb|rgb(r,g,b)|, \verb|wm_render(1)|,
\verb|wm_line(idx,...)|, \verb|wm_clear()|, \verb|screen_w()|/\verb|screen_h()|
(see RotLines.abcl). The shell command \verb|rotlines| is lower-level: it
registers a per-frame draw function via \verb|wm_set_user_render(rot_render)|.

%======================================================================
\section{Sample programs in detail}
At boot, \verb|system/main.c| seeds four files into XFS:
\begin{center}\footnotesize
\begin{tabular}{@{}l l l@{}}
\toprule \textbf{Path} & \textbf{Kind} & \textbf{Focus}\\ \midrule
\verb|/home/hello.c| & C & \verb|printf|\\
\verb|/home/sum.c|   & C & \verb|for| loop\\
\verb|.../PingPong.abcl| & ABCL & inter-actor messaging\\
\verb|.../RotLines.abcl| & ABCL + WM & animation\\
\bottomrule
\end{tabular}
\end{center}

\subsection{hello.c and sum.c}
\begin{code}
int main(void) { printf("hello...\n"); return 0; }
\end{code}
\verb|#include| is accepted/ignored; \verb|printf| is built-in 0; the return
value becomes \verb|aoutRun()|'s result. The simplified bytecode:
\begin{code}
ENTER 0                         ; main has no locals
PUSH_STR offset_of("hello...\n")
CALL_BI 0, 1                    ; printf, 1 arg
PUSH_I32 0
RET
\end{code}
\verb|sum.c| uses a \verb|for| loop (cc expands \verb|for(init;cond;step)| to
\verb|init; while(cond){body; step;}|; note \verb|+= ++ --| are unsupported):
\begin{code}
xsh$ cc /home/sum.c -o /home/sum
xsh$ run /home/sum
sum 1..10 = 55
\end{code}

\subsection{PingPong.abcl --- inter-actor messaging}
\begin{code}
class Player {
    var hits;
    method bounce(other, n) {
        if (n > 0) {
            printf("Player %d: tick (n=%d) hits=%d\n", self, n, hits);
            hits = hits + 1;
            send other.bounce(self, n - 1);
        }
    }
}
main { new Player p1; new Player p2; send p1.bounce(p2, 6); }
\end{code}
\verb|self| is the current actor id (creation order 0,1,\dots); \verb|send| is
asynchronous; when \verb|n-1| reaches 0 no \verb|send| is emitted and the chain
stops:
\begin{code}
Player 0: tick (n=6) hits=0
Player 1: tick (n=5) hits=0
Player 0: tick (n=4) hits=1
...
Player 1: tick (n=1) hits=2
\end{code}

\subsection{RotLines.abcl --- ABCL + WM animation}
\begin{code}
class Rotator {
    var angle;
    method spin(n) {
        if (n > 0) {
            wm_line(0, 512, 384,
                    512 + icos(angle) * 320 / 4096,
                    384 + isin(angle) * 320 / 4096, rgb(255, 80, 80));
            // slots 1..3 at +90/+180/+270 deg likewise
            sleep_ms(16);
            angle = angle + 3;
            send self.spin(n - 1);
        } else { wm_render(0); }
    }
    method start(n) { wm_render(1); send self.spin(n); }
}
main { new Rotator r; send r.start(360); }
\end{code}
\verb|icos|/\verb|isin| return Q12 fixed point (divide by 4096 after scaling by
the radius); \verb|sleep_ms(16)| ≈ 60 fps; sending to \verb|self| forms the
animation loop, and \verb|wm_render(0)| removes the line layer at the end.

\subsection{rotlines (shell command) and apps/abcl\_program.c}
\verb|shell/xsh_rotlines.c| registers \verb|rot_render()| with the WM via
\verb|wm_set_user_render()| and loops advancing the angle with \verb|sleep(16)|;
it draws with \verb|wm_draw_line| (per-frame) rather than the slot-based
\verb|wm_line|. \verb|apps/abcl_program.c| is a sample of the C that abclc
\emph{generates} (a bounded-buffer with \verb|Buffer|/\verb|Producer|/%
\verb|Consumer|/\verb|Controller| classes, mailbox + thread per actor,
\verb|dispatch_*| mega-switch); it shows the correspondence ABCL \verb|class|
\textrightarrow{} C \verb|dispatch_<Class>|, \verb|field| \textrightarrow{}
\verb|enum| index, \verb|send| \textrightarrow{} \verb|enqueue|.

%======================================================================
\section{Development workflow}
\begin{code}
$ ./compile/run-window.sh
xsh$ cd /home && make && ./hello && ./sum
xsh$ edit /home/myprog.c          # C-x C-s save, C-x C-c exit
xsh$ cc myprog.c -o myprog && ./myprog
xsh$ cd /home/abclcp/abclc
xsh$ edit MyActor.abcl && abclc MyActor.abcl && ./MyActor
xsh$ rotlines
\end{code}
Tips: build one target with \verb|make MyActor|; rebuild all with
\verb|rm Makefile && make|; XFS has up to 32768 inodes, so avoid creating many
files.

%======================================================================
\section{Troubleshooting}
\begin{center}\footnotesize
\begin{longtable}{@{}p{5.0cm} p{8.4cm}@{}}
\toprule \textbf{Symptom} & \textbf{Cause / fix}\\ \midrule \endhead
\verb|qemu-system-arm not found| & \verb|brew install qemu| / \verb|apt install qemu-system-arm|\\
stale build boots & \verb|make clean && make PLATFORM=arm-qemu| in \verb|compile/|\\
no shell prompt & \verb|Ctrl-A| \textrightarrow{} \verb|c| for the QEMU monitor, then \verb|info registers|\\
\verb|cc: line N: ...| error & unsupported syntax (structs, pointers, arrays, float, \dots)\\
\verb|abclc: translation failed| & two \verb|send|s in one method, too many \verb|var|s / methods\\
\verb|command not found| but file exists & implicit \verb|run| takes a single path token; use \verb|run PATH ARGS...|\\
no WM window & is it an \verb|arm-qemu| build? did you use \verb|run-window.sh|?\\
files vanished & the RAM disk is volatile --- back up on the host\\
overlong path & \verb|XFS_PATH_MAX = 256|; longer paths are truncated\\
garbled screen & \verb|C-l| in the editor; \verb|clear| in the shell\\
\bottomrule
\end{longtable}
\end{center}

%======================================================================
\section*{Appendix A. File map}
\addcontentsline{toc}{section}{Appendix A. File map}
\begin{center}\footnotesize
\begin{tabular}{@{}l p{9.2cm}@{}}
\toprule \textbf{Area} & \textbf{Key files}\\ \midrule
Boot & \verb|system/main.c|, \verb|system/initialize.c|, \verb|compile/run-*.sh|\\
XFS  & \verb|include/xfs.h|, \verb|system/xfs.c|, \verb|device/ramdisk/|\\
cc   & \verb|include/aout.h|, \verb|system/cc.c|, \verb|system/aout.c|, \verb|shell/xsh_cc.c|\\
abclc& \verb|include/abclc.h|, \verb|system/abclc.c|, \verb|shell/xsh_abclc.c|\\
make/edit & \verb|shell/xsh_make.c|, \verb|shell/xsh_edit.c|\\
WM   & \verb|apps/wm.c|, \verb|apps/abcl_xinu_gui.c|, \verb|shell/xsh_rotlines.c|\\
platform & \verb|compile/platforms/arm-qemu/|, \verb|compile/platforms/arm-rpi/|\\
\bottomrule
\end{tabular}
\end{center}

\section*{Appendix B. Existing samples (after FS seeding)}
\addcontentsline{toc}{section}{Appendix B. Existing samples}
\begin{code}
/home/
├── hello.c
├── sum.c
└── abclcp/
    └── abclc/
        ├── PingPong.abcl
        └── RotLines.abcl
\end{code}
\verb|cd /home && make| builds \verb|hello|, \verb|sum|;
\verb|cd /home/abclcp/abclc && make| builds \verb|PingPong|, \verb|RotLines|.
To go deeper, read \verb|system/cc.c|, \verb|system/aout.c| (the bytecode VM),
and \verb|system/abclc.c| (the ABCL translator).

\end{document}
